---
title: "BUS 235A Final Project"
output: word_document
---

```{r}
library(janitor)
library(dplyr)
library(corrplot)
library(RColorBrewer)
library(caret)
library(ggplot2)
coul <- brewer.pal(5, "Set2")
```

# Loading Data
```{r}
data1 <- read.csv("diamonds.csv", header = TRUE)
```

# Data Cleaning
## 1. Remove empty rows and columns of Data 
```{r}
data2 <- remove_empty(data1, which = c("rows","cols"), quiet = FALSE)

## The argument quiet = FAlSE used for to display message whether there is empty rows/columns in the data1 or not.
```

##2. Remove Duplicate Rows of Data using distinct() function available in dplyr R package
```{r}
diamonds_df <- distinct(data2)

## There is no duplicate data to be removed.
```

# Summarizing Data
```{r}
summary(diamonds_df)
```

## Number of rows and columns
```{r}
cat(sprintf("Number of rows: %d, Number of columns: %d", nrow(diamonds_df),
ncol(diamonds_df)))
```

## Droping column X as we already have the index
```{r}
dF <- subset(diamonds_df, select = -c(X) )
```

```{r}
sapply(dF,class)
```


# Categorical Columns Plots
```{r}
# 1. cut
par(mfrow = c(2,2))
count1 <- table(dF$cut)
barplot(count1, ylab = "count", xlab = "cut", col = coul)

# 2. color
count2 <- table(dF$color)
barplot(count2, ylab = "count", xlab = "color", col = coul)

# 3. clarity
count3 <- table(dF$clarity)
barplot(count3, ylab = "count", xlab = "clarity", col = coul)

```

# Numeric Columns

## Response Variable - Price

```{r}
summary(dF)
```


```{r}
boxplot(dF$price, col = coul, ylab = "price", main = "Boxplot of Price")
```

# scatter plot
```{r}

par(mfrow = c(2,2))

# 1. carat
plot(price ~ carat, data = dF, main = "Price vs carat", col ="blue", pch = 16)
abline(lm(price ~ carat, data = dF))

# 2. depth
plot(price ~ depth, data = dF, main = "Price vs depth", col ="blue", pch = 16)
abline(lm(price ~ depth, data = dF))

# 3. table
plot(price ~ table, data = dF, main = "Price vs table", col ="blue", pch = 16)
abline(lm(price ~ table, data = dF))

# 4. x
plot(price ~ x, data = dF, main = "Price vs x", col ="blue", pch = 16)
abline(lm(price ~ x, data = dF))

# 5. y
plot(price ~ y, data = dF, main = "Price vs y", col ="blue", pch = 16)
abline(lm(price ~ y, data = dF))

# 6. z
plot(price ~ z, data = dF, main = "Price vs z", col ="blue", pch = 16)
abline(lm(price ~ z, data = dF))

```
# VISUALIZING DATA with boxplot
```{r}
# 1. price vs. cut
boxplot(dF$price ~ dF$cut, 
        col = coul,
        border = "brown",
        notch = TRUE,
        ylab = "price", 
        xlab = "carat")


# 2. price vs. color
boxplot(dF$price ~ dF$color, 
        col = coul,
        border = "brown",
        notch = TRUE,
        ylab = "price", 
        xlab = "color")


# 3. price vs. clarity
boxplot(dF$price ~ dF$clarity, 
        col = coul,
        border = "brown",
        notch = TRUE,
        ylab = "price", 
        xlab = "clarity")


```

# Depth and Price
```{r}
# Histogram of Depth
par(mfrow = c(2,2))
hist(dF$depth,xlab = "depth",col = coul,border = "blue")

#  Depth vs Price
plot(price ~ depth, data = dF, main = "Price vs Depth", col ="blue", pch = 16)
abline(lm(price ~ depth, data = dF))

# We can infer from the plot that the Price can vary heavily for the same Depth.

```

# Dimnesion - x, y, z
```{r}
plot(density(dF$x))
lines(density(dF$y))
lines(density(dF$z))
```

Consutruct volume from width, height and length
```{r}
dF$volume <- dF$x * dF$y * dF$z
```

```{r}
# 6. volume
plot(price ~ volume, data = dF, main = "Price vs volume", col ="blue", pch = 16)
abline(lm(price ~ volume, data = dF))
```

## Correlation of numeric data
```{r}
corrplot(cor(dF[,c('price', 'carat','depth','table', 'x','y','z','volume')]), type="full", method = "number" )
```

Looks like price is highly correlated to carat and least correlated to depth.

## factor coversion:
```{r}
# cut quality of the cut (Fair, Good, Very Good, Premium, Ideal)
# color diamond colour, from J (worst) to D (best)
# clarity a measurement of how clear the diamond is (I1 (worst), SI2, SI1, VS2, VS1, VVS2, VVS1, IF (best))
cut_levels <- c("Fair", "Good", "Very Good", "Premium", "Ideal")
clarity_levels <- c("I1", "SI2", "SI1", "VS2", "VS1", "VVS2", "VVS1", "IF")
color_levels <- c("D", "E", "F", "G", "H", "I", "J")
dF$color = factor(dF$color, levels=color_levels)
dF$cut = factor(dF$cut, levels=cut_levels)
dF$clarity = factor(dF$clarity, levels=clarity_levels)
```

## Splitting dataset into 15% test and 85% training data

```{r}
# all variables
set.seed(2)
samples <- sample(1:nrow(dF), size=nrow(dF)/15)
dF.test <- dF[samples, ]
dF.train <- dF[-samples, ]
cat(sprintf("Size of full dataset: %d\n", nrow(dF)))
cat(sprintf("Size of train dataset: %d\n", nrow(dF.train)))
cat(sprintf("Size of test dataset: %d\n", nrow(dF.test)))

```

## Naive model

Model with all the variables
```{r}
model <- lm(price~., data=dF.train)
summary(model)
rmse.train <- sqrt(mean((predict(model, newdata = dF.train) - dF.train$price)^2)) 
rmse.test <- sqrt(mean((predict(model, newdata = dF.test) - dF.test$price)^2)) 
cat(sprintf("r-squared: %.4f  RMSE train: %.2f  RMSE test: %.2f", summary(model)$adj.r.squared, rmse.train, rmse.test))
# 

```



Model without depth & table variable
```{r}
model <- lm(price~carat+cut+color+clarity+x+y+z+volume, data=dF.train)
summary(model)
rmse.train <- sqrt(mean((predict(model, newdata = dF.train) - dF.train$price)^2)) 
rmse.test <- sqrt(mean((predict(model, newdata = dF.test) - dF.test$price)^2)) 
cat(sprintf("r-squared: %.4f  RMSE train: %.2f  RMSE test: %.2f", summary(model)$adj.r.squared, rmse.train, rmse.test))
```


Model without x, y, z variables
```{r}
model <- lm(price~carat+cut+color+clarity+depth+table+volume, data=dF.train)
summary(model)
rmse.train <- sqrt(mean((predict(model, newdata = dF.train) - dF.train$price)^2)) 
rmse.test <- sqrt(mean((predict(model, newdata = dF.test) - dF.test$price)^2)) 
cat(sprintf("r-squared: %.4f  RMSE train: %.2f  RMSE test: %.2f", summary(model)$adj.r.squared, rmse.train, rmse.test))
```


Model without volume variable
```{r}
model <- lm(price~carat+cut+color+clarity+depth+table+x+y+z, data=dF.train)
summary(model)
rmse.train <- sqrt(mean((predict(model, newdata = dF.train) - dF.train$price)^2)) 
rmse.test <- sqrt(mean((predict(model, newdata = dF.test) - dF.test$price)^2)) 
cat(sprintf("r-squared: %.4f  RMSE train: %.2f  RMSE test: %.2f", summary(model)$adj.r.squared, rmse.train, rmse.test))
```


Model with just the carat value
```{r}
model <- lm(price~carat, data=dF.train)
summary(model)
rmse.train <- sqrt(mean((predict(model, newdata = dF.train) - dF.train$price)^2)) 
rmse.test <- sqrt(mean((predict(model, newdata = dF.test) - dF.test$price)^2)) 
cat(sprintf("r-squared: %.4f  RMSE train: %.2f  RMSE test: %.2f", summary(model)$adj.r.squared, rmse.train, rmse.test))
```


Model without table variable
```{r}
model <- lm(price~carat+cut+color+clarity+depth+x+y+z+volume, data=dF.train)
summary(model)
rmse.train <- sqrt(mean((predict(model, newdata = dF.train) - dF.train$price)^2)) 
rmse.test <- sqrt(mean((predict(model, newdata = dF.test) - dF.test$price)^2)) 
cat(sprintf("r-squared: %.4f  RMSE train: %.2f  RMSE test: %.2f", summary(model)$adj.r.squared, rmse.train, rmse.test))
```

Model to predict log of value
```{r}
model <- lm(price~carat, data=dF.train)
summary(model)
rmse.train <- sqrt(mean((predict(model, newdata = dF.train) - dF.train$price)^2)) 
rmse.test <- sqrt(mean((predict(model, newdata = dF.test) - dF.test$price)^2)) 
cat(sprintf("r-squared: %.4f  RMSE train: %.2f  RMSE test: %.2f", summary(model)$adj.r.squared, rmse.train, rmse.test))
```



```{r}
library(DAAG)
cvlm.model <- CVlm(data=dF.train,model, m=4, seed=2, plotit = TRUE, printit = FALSE)
```

```{r}
library(lasso2)
bounds = c(0.001, 0.01, 0.1, 1, 2, 4, 8, 16, 32, 64, 128, 256, 1024, 2048, 4096, 2^13, 2^14, 2^14, 2^15, 2^16)
lasso.models <- l1ce(price~carat+cut+color+clarity+x+y+z+volume, data=dF.train, absolute.t = TRUE, standardize = TRUE, bound = bounds)
cat("Index\t   Bounds\tMSE\n")
for (index in 1:length(bounds)) {
 mse <- sqrt(mean((predict(lasso.models[index]) - dF.train$price)^2))
 cat(sprintf("%4d\t%10.2f\t%.4f\n", index, bounds[index], mse));
}
```


```{r}
#summary(lasso.models[19])
rmse.train <- sqrt(mean((predict(lasso.models[19], newdata = dF.train) - dF.train$price)^2)) 
rmse.test <- sqrt(mean((predict(lasso.models[19], newdata = dF.test) - dF.test$price)^2)) 
cat(sprintf("r-squared: %.4f  RMSE train: %.2f  RMSE test: %.2f", summary(model)$adj.r.squared, rmse.train, rmse.test))
```

